---
title: "Causal Inference with IPTW in riskdiff"
author: "John D. Murphy"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Causal Inference with IPTW in riskdiff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

library(riskdiff)
```

## Introduction

This vignette demonstrates how to use the **riskdiff** package to perform causal inference using Inverse Probability of Treatment Weighting (IPTW). IPTW is a powerful method for estimating causal effects from observational data by creating a pseudo-population where treatment assignment is independent of measured confounders.

### When to Use IPTW

IPTW is particularly useful when:

- You want to estimate **causal effects** (not just associations)
- You have measured all important confounders
- The treatment assignment mechanism is complex
- You want to estimate population-level effects (ATE) or effects on specific subgroups (ATT/ATC)

### Key Assumptions

IPTW relies on three main assumptions:

1. **No unmeasured confounding**: All confounders are measured and included
2. **Positivity**: All subjects have non-zero probability of receiving either treatment
3. **Correct model specification**: The propensity score model is correctly specified

## Basic IPTW Analysis

Let's start with a simple example using the Cachar cancer screening dataset:

```{r basic_example}
# Load the data
data(cachar_sample)

# Quick look at the data
head(cachar_sample)
table(cachar_sample$areca_nut, cachar_sample$abnormal_screen)
```

### Step 1: Calculate IPTW Weights

First, we estimate propensity scores and calculate weights:

```{r calculate_weights}
# Calculate ATE weights for areca nut use
iptw_result <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut", 
  covariates = c("age", "sex", "residence", "smoking", "tobacco_chewing"),
  weight_type = "ATE",
  verbose = TRUE
)

# Examine the results
print(iptw_result)
```

### Step 2: Check IPTW Assumptions

Before interpreting results, we should check key assumptions:

```{r check_assumptions}
# Comprehensive assumption checking
assumptions <- check_iptw_assumptions(iptw_result, verbose = TRUE)
```

### Step 3: Visualize Balance

Visual diagnostics help assess whether weighting achieved balance:

```{r balance_plots, fig.width=8, fig.height=6}
# Create balance plots (requires ggplot2)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plots <- create_balance_plots(iptw_result, plot_type = "both")
  print(plots$love_plot)
  print(plots$ps_plot)
}
```

### Step 4: Estimate Causal Risk Difference

Now we can estimate the causal effect:

```{r causal_effect}
# Estimate ATE using IPTW
rd_causal <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen",
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence", "smoking", "tobacco_chewing"),
  weight_type = "ATE",
  verbose = TRUE
)

print(rd_causal)
summary(rd_causal)
```

## Different Types of Causal Effects

IPTW can estimate different causal estimands depending on the research question:

### Average Treatment Effect (ATE)

The effect of treatment if the entire population received treatment vs. if none received treatment:

```{r ate_example}
rd_ate <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen", 
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence", "smoking"),
  weight_type = "ATE"
)

cat("ATE: The average causal effect of areca nut use in the population\n")
cat("Risk Difference:", scales::percent(rd_ate$rd_iptw, accuracy = 0.01), "\n")
```

### Average Treatment Effect on the Treated (ATT)

The effect among those who actually received treatment:

```{r att_example}
rd_att <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen",
  treatment = "areca_nut", 
  covariates = c("age", "sex", "residence", "smoking"),
  weight_type = "ATT"
)

cat("ATT: The average causal effect among areca nut users\n")
cat("Risk Difference:", scales::percent(rd_att$rd_iptw, accuracy = 0.01), "\n")
```

### Average Treatment Effect on the Controls (ATC)

The effect among those who did not receive treatment:

```{r atc_example}
rd_atc <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen",
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence", "smoking"), 
  weight_type = "ATC"
)

cat("ATC: The average causal effect among non-users of areca nut\n")
cat("Risk Difference:", scales::percent(rd_atc$rd_iptw, accuracy = 0.01), "\n")
```

## Advanced Options

### Bootstrap Confidence Intervals

For small samples or when assumptions are questionable, bootstrap confidence intervals may be more robust:

```{r bootstrap_example}
rd_bootstrap <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "head_neck_abnormal",
  treatment = "tobacco_chewing",
  covariates = c("age", "sex", "residence", "areca_nut"),
  bootstrap_ci = TRUE,
  boot_n = 500,  # Use more in practice (1000+)
  verbose = FALSE
)

print(rd_bootstrap)
```

### Different Propensity Score Models

You can specify different link functions for the propensity score model:

```{r ps_models}
# Logistic regression (default)
ps_logit <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "tobacco_chewing",
  covariates = c("age", "sex", "residence", "areca_nut"),
  method = "logistic"
)

# Probit regression
ps_probit <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "tobacco_chewing", 
  covariates = c("age", "sex", "residence", "areca_nut"),
  method = "probit"
)

# Compare propensity score distributions
cat("Logistic PS range:", round(range(ps_logit$ps), 3), "\n")
cat("Probit PS range:", round(range(ps_probit$ps), 3), "\n")
```

### Weight Stabilization and Trimming

Stabilized weights often have better properties:

```{r stabilization}
# Unstabilized weights
ps_unstab <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence"),
  stabilize = FALSE
)

# Stabilized weights (default)
ps_stab <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence"), 
  stabilize = TRUE
)

cat("Unstabilized weight variance:", round(var(ps_unstab$weights), 2), "\n")
cat("Stabilized weight variance:", round(var(ps_stab$weights), 2), "\n")
```

Extreme weights can be problematic and may need trimming:

```{r trimming}
# Check for extreme weights
summary(ps_stab$weights)

# Trim at 1st and 99th percentiles
ps_trimmed <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence"),
  trim_weights = TRUE,
  trim_quantiles = c(0.01, 0.99)
)

cat("Original weight range:", round(range(ps_stab$weights), 2), "\n")
cat("Trimmed weight range:", round(range(ps_trimmed$weights), 2), "\n")
```
  treatment = "smoking",
  covariates = c("maternal_age", "race", "education"),
  trim_weights = TRUE,
  trim_quantiles = c(0.01, 0.99)
)

cat("Original weight range:", round(range(ps_stab$weights), 2), "\n")
cat("Trimmed weight range:", round(range(ps_trimmed$weights), 2), "\n")
```

## Comparison with Traditional Regression

Let's compare IPTW results with traditional regression adjustment:

```{r comparison}
# Traditional regression-based risk difference
rd_regression <- calc_risk_diff(
  data = cachar_sample,
  outcome = "abnormal_screen",
  exposure = "areca_nut",
  adjust_vars = c("age", "sex", "residence", "smoking"),
  link = "auto"
)

# IPTW-based causal risk difference  
rd_iptw <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen",
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence", "smoking"),
  weight_type = "ATE"
)

# Compare results
comparison_table <- data.frame(
  Method = c("Regression Adjustment", "IPTW (ATE)"),
  Risk_Difference = scales::percent(c(rd_regression$rd, rd_iptw$rd_iptw), accuracy = 0.01),
  CI_Lower = scales::percent(c(rd_regression$ci_lower, rd_iptw$ci_lower), accuracy = 0.01),
  CI_Upper = scales::percent(c(rd_regression$ci_upper, rd_iptw$ci_upper), accuracy = 0.01),
  P_Value = sprintf("%.3f", c(rd_regression$p_value, rd_iptw$p_value))
)

print(comparison_table)
```

## Troubleshooting Common Issues

### Poor Balance

If covariates remain imbalanced after weighting:

```{r poor_balance, eval=FALSE}
# Check which variables have poor balance
assumptions <- check_iptw_assumptions(iptw_result)
poor_balance_vars <- assumptions$balance$poor_balance_vars

if (length(poor_balance_vars) > 0) {
  cat("Variables with poor balance:", paste(poor_balance_vars, collapse = ", "), "\n")
  
  # Try including interactions or polynomial terms
  iptw_improved <- calc_iptw_weights(
    data = cachar_sample,
    treatment = "areca_nut",
    covariates = c("age", "I(age^2)", "sex", "residence", 
                   "smoking", "age:sex"),  # Add interactions
    weight_type = "ATE"
  )
}
```

### Extreme Propensity Scores

When subjects have very high or low propensity scores:

```{r extreme_ps, eval=FALSE}
# Check propensity score distribution
iptw_result <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence")
)

# Identify subjects with extreme scores
extreme_low <- which(iptw_result$ps < 0.05)
extreme_high <- which(iptw_result$ps > 0.95)

if (length(extreme_low) > 0 || length(extreme_high) > 0) {
  cat("Consider trimming sample to region of common support\n")
  
  # Restrict to common support
  common_support <- iptw_result$ps >= 0.05 & iptw_result$ps <= 0.95
  data_restricted <- cachar_sample[common_support, ]
  
  # Re-analyze with restricted sample
  rd_restricted <- calc_risk_diff_iptw(
    data = data_restricted,
    outcome = "abnormal_screen",
    treatment = "areca_nut", 
    covariates = c("age", "sex", "residence")
  )
}
```

### Model Specification

Testing different model specifications:

```{r model_spec, eval=FALSE}
# Simple model
ps_simple <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "sex")
)

# Complex model with interactions
ps_complex <- calc_iptw_weights(
  data = cachar_sample,
  treatment = "areca_nut",
  covariates = c("age", "I(age^2)", "sex", "residence", 
                 "smoking", "tobacco_chewing", "age:sex")
)

# Compare balance
check_iptw_assumptions(ps_simple, verbose = FALSE)
check_iptw_assumptions(ps_complex, verbose = FALSE)
```

## Sensitivity Analysis

IPTW assumes no unmeasured confounding. Sensitivity analysis can assess robustness:

```{r sensitivity, eval=FALSE}
# Simulate an unmeasured confounder
set.seed(123)
cachar_sample$unmeasured_confounder <- rbinom(nrow(cachar_sample), 1, 0.3)

# Compare results with and without the unmeasured confounder
rd_without_u <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen",
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence")
)

rd_with_u <- calc_risk_diff_iptw(
  data = cachar_sample,
  outcome = "abnormal_screen", 
  treatment = "areca_nut",
  covariates = c("age", "sex", "residence", "unmeasured_confounder")
)

cat("Without unmeasured confounder:", scales::percent(rd_without_u$rd_iptw), "\n")
cat("With unmeasured confounder:", scales::percent(rd_with_u$rd_iptw), "\n")
cat("Difference:", scales::percent(abs(rd_without_u$rd_iptw - rd_with_u$rd_iptw)), "\n")
```

## Best Practices

1. **Always check assumptions** before interpreting results
2. **Visualize balance** using love plots and propensity score distributions  
3. **Consider the estimand** (ATE vs. ATT vs. ATC) based on your research question
4. **Use stabilized weights** unless you have a specific reason not to
5. **Trim extreme weights** cautiously - they may indicate model misspecification
6. **Compare with regression adjustment** as a sensitivity check
7. **Report effective sample size** to assess precision loss
8. **Consider bootstrap CIs** for small samples or non-normal distributions

## Reporting IPTW Results

When reporting IPTW analyses, include:

1. **Propensity score model specification** and diagnostics
2. **Balance assessment** before and after weighting
3. **Weight distribution** and any trimming performed
4. **Effective sample size** and precision considerations
5. **Sensitivity analyses** when possible
6. **Clear statement of estimand** (ATE/ATT/ATC)

## Conclusion

IPTW provides a powerful framework for causal inference from observational data. The **riskdiff** package makes these methods accessible while providing essential diagnostics and visualizations. Remember that causal inference requires careful thought about study design, confounders, and assumptions - the methods are only as good as these foundational elements.

For more advanced applications, consider methods like:
- Marginal structural models for time-varying treatments
- Doubly robust estimation combining IPTW with outcome modeling
- Machine learning approaches for propensity score estimation

## References

Austin, P. C. (2011). An introduction to propensity score methods for reducing the effects of confounding in observational studies. *Multivariate Behavioral Research*, 46(3), 399-424.

Hernán, M. A., & Robins, J. M. (2020). *Causal inference: What if*. Boca Raton: Chapman & Hall/CRC.

Robins, J. M., Hernán, M. A., & Brumback, B. (2000). Marginal structural models and causal inference in epidemiology. *Epidemiology*, 11(5), 550-560.
