---
title: "Getting Started with Risk Differences"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Getting Started with Risk Differences}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>",
 fig.width = 7,
 fig.height = 4,
 fig.align = "center"
)
```

```{r load libraries}
library(riskdiff)
library(dplyr)
library(ggplot2)
```

## Why Risk Differences Matter in Public Health

When communicating health risks to policymakers, patients, or the public, **absolute measures** like risk differences are often more meaningful than **relative measures** like odds ratios or risk ratios.

Consider these two statements about betel nut (areca nut) chewing and cancer risk:

1.  "Betel nut chewing increases the odds of cancer by 5.2 times"

2.  "Betel nut chewing increases cancer risk by 12 percentage points"

The second statement is immediately actionable: in a screening group of 1,000 people, you would expect to find approximately 120 additional cancer cases among betel nut chewers compared to non-chewers. This directly informs:

-   Resource allocation for screening programs

-   Expected yield from targeted interventions

-   Number needed to screen calculations

-   Public health messaging priorities

> **Key Concept**: Risk differences represent the **additional burden of disease** attributable to an exposure. They are on the same scale as the outcome, making them intuitive for non-statistical audiences.

## Understanding the Example Data

The `riskdiff` package includes example data from a cancer screening program in Northeast India:

```{r load example data}
data(cachar_sample)

# Basic structure
glimpse(cachar_sample)

# Key variables for our analysis
cachar_sample %>%
  select(abnormal_screen, areca_nut, smoking, alcohol, age, sex, residence) %>%
  summary()
```

This dataset represents a cross-sectional screening study where:

-   `abnormal_screen` = 1 indicates an abnormal cancer screening result

-   `areca_nut` indicates areca (or betel) nut chewing status

-   Other variables capture demographics and risk factors

## Your First Risk Difference Analysis

Let's calculate the risk difference for cancer associated with betel nut chewing:

```{r simple unadjusted analysis}
# Simple unadjusted analysis
rd_simple <- calc_risk_diff(
  data = cachar_sample,
  outcome = "abnormal_screen",
  exposure = "areca_nut"
)

print(rd_simple)
```

### Understanding the Output

The output shows:

-   **rd**: The risk difference (e.g., 0.082 = 8.2 percentage points)

-   **ci_lower, ci_upper**: 95% confidence interval bounds

-   **p_value**: Test of whether the risk difference equals zero

-   **model_type**: Which GLM link function successfully converged

-   **n_obs**: Number of observations used in the analysis

### Interpreting Risk Differences

```{r visualisation}
# Let's visualize what this risk difference means
exposure_summary <- cachar_sample %>%
  group_by(areca_nut) %>%
  summarise(
    n = n(),
    cases = sum(abnormal_screen),
    risk = mean(abnormal_screen),
    se = sqrt(risk * (1 - risk) / n)
  ) %>%
  mutate(
    risk_percent = risk * 100,
    se_percent = se * 100
  )

print(exposure_summary)

# The risk difference is simply:
rd_value <- diff(exposure_summary$risk)
cat("Risk difference:", round(rd_value * 100, 1), "percentage points\n")
```

### Adjusted Analyses

Real-world associations are often confounded. Let's adjust for age and sex:

```{r adjusted analysis}
# Age and sex adjusted analysis
rd_adjusted <- calc_risk_diff(
  data = cachar_sample,
  outcome = "abnormal_screen",
  exposure = "areca_nut",
  adjust_vars = c("age", "sex")
)
print(rd_adjusted)

# Show comparison in a readable format
cat("\n=== COMPARISON: UNADJUSTED vs ADJUSTED ===\n")
cat("Unadjusted Risk Difference:", sprintf("%.2f%%", rd_simple$rd * 100), 
    sprintf("(%.2f%%, %.2f%%)", rd_simple$ci_lower * 100, rd_simple$ci_upper * 100), "\n")
cat("Adjusted Risk Difference:  ", sprintf("%.2f%%", rd_adjusted$rd * 100), 
    sprintf("(%.2f%%, %.2f%%)", rd_adjusted$ci_lower * 100, rd_adjusted$ci_upper * 100), "\n")
cat("Difference in estimates:   ", sprintf("%.2f%%", (rd_adjusted$rd - rd_simple$rd) * 100), "percentage points\n")
```

> **Note**: Adjustment often changes the risk difference estimate. This indicates that age and/or sex were confounders of the chewing-cancer association.

### Stratified Analysis

Sometimes we want to know if effects differ across subgroups:

```{r stratified analysis}
# Stratified by residence (urban vs rural)
rd_stratified <- calc_risk_diff(
  data = cachar_sample,
  outcome = "abnormal_screen",
  exposure = "areca_nut",
  strata = "residence"
)

print(rd_stratified)
```

This shows separate risk differences for urban and rural areas, which might reflect:

-   Different exposure patterns
-   Access to healthcare
-   Environmental co-factors

### Visualizing Your Results

A picture is worth a thousand p-values:

```{r visualise stratified results}

# Stratified analysis by residence
rd_stratified <- calc_risk_diff(
  data = cachar_sample,
  outcome = "abnormal_screen",
  exposure = "areca_nut",
  strata = "residence"
)

print(rd_stratified)

# Check if estimates are stable enough for visualization
rd_summary <- rd_stratified %>%
  summarise(
    n_valid = sum(!is.na(rd)),
    max_ci_width = max((ci_upper - ci_lower) * 100, na.rm = TRUE),
    .groups = "drop"
  )

# Only create plot if we have reasonable estimates
if (rd_summary$n_valid > 0 && rd_summary$max_ci_width < 200) {
  # Use the fixed forest plot function
  plot_obj <- create_forest_plot(
    rd_stratified, 
    title = "Risk Difference for Cancer by Areca Nut Chewing",
    max_ci_width = 30
  )
  
  if (!is.null(plot_obj)) {
    print(plot_obj)
  } else {
    cat(paste0(riskdiff:::.safe_warning()), "Could not create plot due to unstable estimates.\n")
  }
} else {
  cat(paste0(riskdiff:::.safe_warning()), "Estimates too unstable for meaningful visualization.\n")
  cat("Consider:\n")
  cat(paste0(riskdiff:::.safe_bullet()), "Larger sample sizes\n") 
  cat(paste0(riskdiff:::.safe_bullet()), "Different stratification variables\n")
  cat(paste0(riskdiff:::.safe_bullet()), "Pooled analysis instead of stratification\n")
  
  # Show tabular results instead using the robust summary function
  formatted_results <- create_summary_table(
    rd_stratified, 
    caption = "Risk Differences by Residence"
  )
  
  knitr::kable(formatted_results, caption = "Risk Differences by Residence")
}
```

### Common Pitfalls and Solutions

#### 1. Model Convergence Issues

The identity link GLM (which directly estimates risk differences) often fails to converge. The `riskdiff` package handles this automatically:

```{r convergence handling}
# Force different link functions
rd_identity <- calc_risk_diff(
  cachar_sample, "abnormal_screen", "areca_nut", 
  link = "identity"
)

rd_log <- calc_risk_diff(
  cachar_sample, "abnormal_screen", "areca_nut", 
  link = "log"
)

# Compare model types used
cat("Identity link model type:", rd_identity$model_type, "\n")
cat("Log link model type:", rd_log$model_type, "\n")
```

#### 2. Very Rare Outcomes

With rare outcomes, risk differences become very small:

```{r rare outcomes}
# Create a rare outcome (1% prevalence)
cachar_sample$rare_outcome <- rbinom(nrow(cachar_sample), 1, 0.01)

rd_rare <- calc_risk_diff(
  cachar_sample, 
  "rare_outcome", 
  "areca_nut"
)

print(rd_rare)
```

> **Tip**: For very rare outcomes (\<1%), consider whether risk ratios might be more appropriate for your research question.

#### 3. Missing Data

The package automatically handles missing data by complete case analysis:

```{r missing data}
# Create a copy with some missing data for demonstration
set.seed(123)  # For reproducibility
cachar_with_missing <- cachar_sample %>%
  mutate(
    # Introduce more modest missing data (~3% in age, ~2% in alcohol)
    age_with_missing = ifelse(runif(n()) < 0.03, NA, age),
    alcohol_with_missing = ifelse(runif(n()) < 0.02, NA, alcohol)
  )

# Check the missing data patterns
missing_summary <- cachar_with_missing %>%
  summarise(
    total_observations = n(),
    age_missing = sum(is.na(age_with_missing)),
    alcohol_missing = sum(is.na(alcohol_with_missing)),
    total_missing_any = sum(!complete.cases(select(., age_with_missing, alcohol_with_missing, abnormal_screen, areca_nut))),
    complete_cases = sum(complete.cases(select(., age_with_missing, alcohol_with_missing, abnormal_screen, areca_nut)))
  )

print(missing_summary)

# Analysis with variables that have missing data
rd_missing <- calc_risk_diff(
  cachar_with_missing,
  "abnormal_screen",
  "areca_nut",
  adjust_vars = c("age_with_missing", "alcohol_with_missing")
)

# Compare with complete case analysis
rd_complete <- calc_risk_diff(
  cachar_sample,
  "abnormal_screen", 
  "areca_nut",
  adjust_vars = c("age", "alcohol")
)

cat("\n=== IMPACT OF MISSING DATA ===\n")
cat("Complete data analysis (n=", rd_complete$n_obs, "): ", sprintf("%.2f%%", rd_complete$rd * 100), 
    sprintf(" (%.2f%%, %.2f%%)", rd_complete$ci_lower * 100, rd_complete$ci_upper * 100), "\n")

# Check if missing data analysis succeeded
if (!is.na(rd_missing$rd)) {
  cat("Missing data analysis (n=", rd_missing$n_obs, "):  ", sprintf("%.2f%%", rd_missing$rd * 100), 
      sprintf(" (%.2f%%, %.2f%%)", rd_missing$ci_lower * 100, rd_missing$ci_upper * 100), "\n")
  cat("Cases lost to missing data: ", rd_complete$n_obs - rd_missing$n_obs, "\n")
} else {
  cat("Missing data analysis: FAILED (insufficient data or convergence issues)\n")
  cat("Attempted to use n =", rd_missing$n_obs, "complete cases\n")
  cat("Cases lost to missing data: ", nrow(cachar_with_missing) - rd_missing$n_obs, "\n\n")
  
  cat("LESSON: This demonstrates why missing data can be problematic:\n")
  cat(paste0(riskdiff:::.safe_bullet()), "Listwise deletion can dramatically reduce sample size\n")
  cat(paste0(riskdiff:::.safe_bullet()), "Small samples may cause model convergence failures\n") 
  cat(paste0(riskdiff:::.safe_bullet()), "Consider multiple imputation for better missing data handling\n")
  cat(paste0(riskdiff:::.safe_bullet()), "The riskdiff package gracefully handles these failures\n")
}
```

### Quick Reference

#### Basic Syntax

```{r basic syntax example, eval=FALSE}
# Example usage:
result <- calc_risk_diff(
  data = cachar_sample,           # Your dataset
  outcome = "abnormal_screen",    # Binary outcome variable (0/1)
  exposure = "areca_nut",         # Exposure of interest
  adjust_vars = c("age", "sex"),  # Variables to adjust for
  strata = "residence",           # Stratification variables
  link = "auto",                  # Link function: "auto", "identity", "log", "logit"
  alpha = 0.05,                   # Significance level (0.05 = 95% CI)
  verbose = FALSE                 # Print diagnostic messages if TRUE
)
```

#### Interpretation Guide

| Risk Difference | Interpretation | Public Health Meaning |
|------------------------|------------------------|------------------------|
| 0.05 (5%) | 5 percentage point increase | 50 extra cases per 1,000 screened |
| 0.01 (1%) | 1 percentage point increase | 10 extra cases per 1,000 screened |
| -0.03 (-3%) | 3 percentage point decrease | 30 fewer cases per 1,000 screened |

### When to Use Risk Differences

✅ **Use risk differences when:**

-   Communicating to non-statistical audiences

-   Making policy decisions about interventions

-   The outcome is relatively common (\>5%)

-   You need to calculate number needed to treat/screen

-   Comparing absolute impact across different populations

❌ **Consider alternatives when:**

-   Outcome is very rare (\<1%)

-   You need to compare across populations with very different baseline risks

-   Your research question is about biological/etiological mechanisms

## Next Steps

Now that you understand the basics:

1.  See the **"Complete Example"** vignette for a full analysis workflow

2.  Check the **"Technical Details"** vignette for statistical methodology

3.  Use `?calc_risk_diff` for detailed function documentation

## Getting Help

-   **Issues or bugs**: [https://github.com/jackmurphy2351/riskdiff/issues](https://github.com/jackmurphy2351/riskdiff)

-   **Function help**: `?calc_risk_diff`, `?format_risk_diff`

-   **All vignettes**: `browseVignettes("riskdiff")`

------------------------------------------------------------------------

*This vignette is part of the riskdiff package (v0.2.0), developed to make risk difference calculations accessible to public health researchers.*
